---
title: 'An Exploratory Analysis of Fertility and Schooling'
author: "Paul Laskowski"
date: "August 15, 2016"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This analysis is motivated by the following research question:

\begin{center}
How is the amount of schooling received by a woman related to the number of children she has?
\end{center}

We will address this question using exploratory data analysis techniques. Our data comes from the General Social Survey (GSS), a 
large-scale sociodemographic study that has taken place since 1972.
Nowadays, the GSS takes place every other year.  The survey is 
conducted face-to-face and takes about 90 minutes to administer.

Our procedure is loosely based on an analysis by Sander.\footnote{Sander. The Effect of Women’s Schooling on Fertility. \textit{Economics Letters} 
40, 229-233}  Unlike Sander's study, we will not perform any
statistical modeling, but we will dive deeper into an exploratory
analysis of the data.

We will use a pooled cross section of data, including the years 2010,
2012, and 2014.

In line with Sander, we will restrict our analysis to women
between the ages of 35 and 44.  The reason for the lower bound is 
that most women age 35 and above have reached the point at which 
they will not have more children.  This is convenient for our 
analysis, because we don't have to worry if participants 
with few years of education are simply so young that they haven't 
had all of the children they're going to have.  Similarly, most
women above age 34 have completed their education, which also 
simplifies our analysis.

The upper bound on age serves a different purpose.  We want to 
compare women born in roughly the same time period.  Over the 
last decades, we know that the average amount of education has
grown, while the average number of children has decreased.  As a 
result, women with a high amount of education tend to be born in 
more recent years, and for this reason alone, they will also tend 
to to have fewer children.  By capping the age of women in our 
sample at 44, we reduce the possibility that age drives the
relationships we observe.


Some caveats to keep in mind:

1. The GSS is a survey, and our sample is not representative of the
U.S. population.  Statisticians have specialized techniques for 
working with surveys.  For example, the General Social Survey
assigns a weight to each observation, which is meant to account for
differences in how likely each individual is to appear in the sample.
The current analysis is limited to descriptive statistics, so we
will only be making claims about the sample, not the population.

2. Our analysis is not causal.  The techniques we use here
do not address the question of whether more education causes
people to have more or less children.  Note that Sander's paper 
claims to measure the causal effect of education, but there are 
good reasons to be suspicious of this claim.  We will return 
to issues of causality later in the course.

## Setup

First, we load the car library, which gives us a convenient scatterplotMatrix function.

```{r car}
library(car)
```


I used the GSS data explorer (https://gssdataexplorer.norc.org)
to download a dataset with just the variables are are interested in.
I performed some minimal editing to make sure the variable names
were correct, and saved the data to the file Fertility.RData.

When loading this file, we don't need to set the working directory, because we put the data file in the same directory as our Rmd file.  A quirk of Knitr is that the working directory is reset to the place the Rmd file is located with every new chunk (we could override this with a command like opts_knit\$set(root.dir = '~/Desktop/data_w203')).  To keep things simple, we try to always keep our data together with our script files.

```{r loading}
# Load the data 
load("Fertility.Rdata")
```

## Data Selection

We note that we have `r nrow(Fert_all)` observations and `r length(colnames(Fert_all))` variables.

```{r shape}
nrow(Fert_all)
str(Fert_all)
```

Looking at the age variable and the sex variable, we notice that
we have both men and women and individuals as young as 18.

```{r age_range}
summary(Fert_all$age)
summary(Fert_all$sex)
```

Our first step is to subset the data.  To do this we will select
women that are at least 35 years old using a logical vector.

```{r}
subcases = 35 <= Fert_all$age & Fert_all$age < 45 & Fert_all$sex == "Female" & 
  !is.na(Fert_all$age) & ! is.na(Fert_all$sex)
```

Note that we have to specifically check if age or sex are NA.  If
we don't do this, observations for which age or sex are missing
will show up as NA's in our logical vector and we will end up keeping
them.  Since we want to ensure that all our individuals belong to the 
subgroup of women aged 35 and above, we use the is.na function to
make sure such cases end up as False in our logical vector.

We use our logical vector to pull out just the rows we want and 
save them into a new data frame.

```{r}
Fert = Fert_all[subcases, ]
nrow(Fert)
```

We could have accomplished the same operation using the convenient
subset command

```{r}
Fert = subset(Fert_all, sex == "Female" & 35 <= age & age < 45)
nrow(Fert)
```

Our new data frame has just `r nrow(Fert)` observations.  We examine a summary.

```{r}
summary(Fert)
```

Looking at our subset of data, we note that there are some missing
values.  We don't know the number of kids for 4 individuals, and
we don't have years of education for 1.  This is a reasonably small
fraction of our cases.

## Exploratory Analysis

When we have a large number of variables, we often
begin with a scatterplot matrix.  This is helpful
for getting a high-level overview of the relationships
between our variables and can draw our attention
to important features we want to investigate further.

```{r}
scatterplotMatrix( ~ educ + kids + age + year, data = Fert,
                   main = "Scatterplot Matrix for Key GSS Variables")
```

We notice a couple of features in this matrix that 
can help guide our analysis.

1. There is a noticable negative relationship between
educ and kids.  This is the relationship we are most
interested in.

2. We notice that age and year have much weaker relationships
with the other variables.  This is good, because these
variables could confound our analysis.  We can pick out
some interesting features.  The educ vs. age plot reveals
a noticable downward tick for low levels of education.

3.  Year does not seem to have a strong relationship with
any other variable, which is good since
we're pooling three years together.  There seems to be an
effect where women with the most kids are sampled in the
later years.

This plot suggests that educ and kids are closely related and
it also highlights that we should worry that age (and to a lesser
extent year) could affect the bivariate relationships we see.

Though we will not do any modeling, we will think of kids as our
outcome variable.  We first summarize it.

```{r}
summary(Fert$kids)
```

We notice that kids seems to take on integer, nonnegative values,
as we would expect.  Next, we create a histogram of kids, 
making sure that we set the  cut points correctly using 
the breaks argument.  We also supply tick marks on the 
x-axis manually.

```{r}

hist(Fert$kids, breaks = 0:8 - 0.5, main = "Number of Children", 
     xlab = NULL)
axis(1, at = 0:7)
```

We notice a few important features here:

1. kids takes on integer values.  Moreover, the maximum is
only 7, so integer effects may play an important role.  That is,
a model that assumes kids to be continuous may be distoring.

2. kids is also nonnegative, and this boundary seems important
in the sense that there is a sharp dropoff at zero.  It
is as though the left end of a larger distribution is being
chopped off.  kids is what we would call a count variable.

3. Visually, the histogram seems to have a positive skew
(right skew).  This means that there are observations stretching
further to the right of the bulk of the data.  We note that 
the mean is greater than the median, which is typically 
what we see for positively skewed variables.

Because kids is a count variable, we may need to consider special
techniques when we model it.  For example, there is an advanced 
technique called Poisson regression that may be applicable here.

We next examine years of education, which we consider to be
our main input variable.  As before, we must manually set the
cut points for our histogram.

```{r}
summary(Fert$educ)
hist(Fert$educ, breaks = -1:20 + 0.5, main = "Years of Education", 
     xlab = NULL)
```

We notice several features from the summary and histogram:

1. Like kids, education seems to take on integer values.

2. Unlike kids, years of education seems more dispersed, so
integer effects are not as important.  Additionally, the variable
is non-negative, but the lower boundary does not seem important.
The vast majority of the data is greater than zero.

3. There is an unusual spike at 12 years of education.  This 
corresponds to individuals that finish high school but do not
go on to college.  Similarly, a smaller spike can be discerned
at 16 years, which corresponds to individuals that finish a 4-year
college.  This is important to note because there may be something
special about graduating high school, or graduating college, as 
opposed to getting an extra year of education that does not result
in graduation.  Later in the course, we will learn about index
variables, which are a technique we can apply to model such
graduation effects.  For the current analysis, we will remember
to check for graduation effects as we continue exploring.

We next examine our age variable.

```{r}
summary(Fert$age)
hist(Fert$age, breaks = 34:44 + 0.5, main = "Age of Participants",
     xlab = "years")
axis(1, at = 35:44)
```

Note that age ranges from 35 to 44 and the distribution seems
uniform in this range.

We want to understand what bivariate relationship exists
between our main variables of interest, kids and educ.  We begin with a scatterplot.  Because each variable takes on a rather limited set of
values, we add jitter along each axis to make sure points don't 
overlap perfectly.  We also add the ordinary least squares 
regression line, which is a way of visualizing what linear 
relationship exists in the data.  We will have much more to
say about this line later in the course.

```{r}
plot(jitter(Fert$educ, factor=2), jitter(Fert$kids, factor=2),  
     xlab = "education", ylab = "children", 
     main = "Number of Children for Different Levels of Education")
abline(lm(Fert$kids ~ Fert$educ))
```

Without going into too much detail, this plot tells us that
there is an overall negative linear relationship between our
variables.  More education is associated with less children,
on the average.  We can confirm this by checking the sample
correlation

```{r}
cor(Fert$kids, Fert$educ, use = "complete.obs")
```

Note that the cor function does not accept an na.rm argument.
instead, we set the use argument to "complete.obs", which tells
the function to only look at rows for which all variables have
a value.

The correlation of `r cor(Fert$kids, Fert$educ, use = "complete.obs")` is indeed negative, though it is moderate in magnitude.

Regression is a useful tool, but this plot only tells us what
the best fitting line looks like, it does not reveal more
complicated relationships, or whether the relationship between
our variables is actually linear.

To gain a more detailed view, we try using a boxplot.  We use the
at argument to specify where the boxes go along the x-axis.
If we didn't do this, the box corresponding to 0 years of 
education would be right next to the box for 3 years, since
there are no data points in between.  By putting the unique
values of education into the at argument, we ensure that the
boxes are spaced linearly.

We also use the cex.axis argument to scale the numbers on the
axes, to prevent them from overlapping and being cut off.

```{r}
boxplot(kids ~ educ, data = Fert,
        at = sort(unique(Fert$educ)), cex.axis = .7,
        main = "Number of Children by Years of Education",
        xlab = "education", ylab = "children")
```

Overall, the relationship does not appear especially linear.
We shouldn't pay too much attention to the left 3 or 4 bars
of the graph, since there are so few data points there.
Up to about 11 years, the trend seems to be flat, or perhaps
even increasing.  At 12 years, the number of kids drops and
the trend to the right of this point seems decreasing.
 
To better assess this relationship, we will plot the mean 
number of kids for each level of education.  We can use a 
by statement to get the means.

```{r}
kid_means = by(Fert$kids, Fert$educ,  mean, na.rm = T)
```

To plot the means correctly, we must enter the x-values manually
into the plot command.  We can use the unique function to extract
the x-values and the sort function to sort them so that they work
with the plot command.

```{r}
plot(sort(unique(Fert$educ)), kid_means, 
     xlab = "education", ylab = "children", 
     main = "Mean Number of Children by Years of Education")
```

There is a lot of noise on the left side of the graph, and 0 years of
education appears as a notable outlier.  We might wonder if 
respondents that entered 0 misunderstood the question, or if 
their answers were recorded incorrectly.  Setting this point aside,
the overall trend could be approximated by a parabola.  Later in 
the course, we will discuss the use of quadratic terms to model 
parabolic relationships.

We mentioned the possibility of a graduation effect at
12 years and 16 years of education, and we can see some evidence
for this in the box plot and the plot of means.  The mean at 12 
years of education seems noticably lower than the trend left
of that point.  A similar downward jog can be seen at 16 years
of education.

We might speculate that participants that graduate from high
school or college might be especially career-focused, resulting
in the lower number of children.  Given the variables we have
in this dataset, we are not able to futher investigate this
possibility.

As we mentioned before, we can use a technique called indicator
variables to model these types of discontinuities.  We will learn
about indicator variables later in the course and discuss the
tradeoffs inherent in choosing a more complicated model.

To focus our attention on graduation events, we may want to bin
our educ variable into the intervals between graduations. This
is easily done with the cut command.

```{r}
educ_bin = cut(Fert$educ, breaks = c(0,11,15, Inf), labels = 
      c("Some Primary School", "High School Graduate", "College Graduate"))
summary(educ_bin)
```

The resulting boxplot then clearly shows the 
number of kids for each group, telling a different kind
of story.

```{r}
boxplot(kids ~ educ_bin, data = Fert, cex.axis = .9,
        main = "Number of Children by Educational Attainment",
        ylab = "children")
```

Finally, we want to examine how our age variable relates to 
education and to kids.  This will help us assess whether the 
effects we observe could be driven by age, and whether our
strategy of limiting our analysis to women between the ages
of 35 and 44 was successful.  

First we examine the relationship between age and educ.
We could use a boxplot or a plot of means.  Here, we select
a plot of means.

```{r}
age_means = by(Fert$age, Fert$educ,  mean, na.rm = T)
plot(sort(unique(Fert$educ)), age_means, 
     xlab = "education", ylab = "age", 
     main = "Mean Age by Years of Education")
```

We observe that women with less than 8 years of education
appear to be several years younger than women with at least 
8 years.  Above 8 years of education, women in each column 
seem to have roughly the same distribution of ages, as we 
hoped to achieve.

We next examine the relationship between age and kids.  Again,
we select a plot of means.

```{r}
kids_by_age = by(Fert$kids, Fert$age,  mean, na.rm = T)
plot(sort(unique(Fert$age)), kids_by_age, 
     xlab = "age", ylab = "kids", 
     main = "Mean Number of Children by Age")
```

In this graph, we don't see a very strong relationship
between age and number of kids.  There is a noticable
outlier at 44 years of age.  In particular, we are curious
to see if the younger women in the sample have fewer kids,
which might explain why the number of kids we saw for low
levels of education was so low.  However, women on the 
left side of this graph do not seem to stand out.  Note
that the range on the y-axis is less than 0.5 children,
so this plot does not really explain the larger differences
we saw earlier.

Even though we don't see a clear story about how age could
influence the results we see, the low age of women
with 1-7 years of education does suggest that this group
of women might be especially unique in some way.  This is
especially plausible given the small number of women in this
group and the fact that the US school system typically requires
students to remain in school until the age of 16, when most
would have at least 10 years of education.

Above 7 years of education, we don't see any strong relationship
between age and our other variables, which strengthens our
belief that the women in this group can be meaningfully
compared against each other.

## Discussion

Our exploratory analysis has uncovered a number of features
that would prove useful as we turn to statistical modeling.
Among the important features we found, there were skewed 
distributions, attenuated / count variables, (possibly quadratic) 
non-linearities, outliers, and graduation effects.
Later, we will learn more about statistical modeling, and we will use
some of these same exploratory techniques to test the assumptions
of our models.