%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Statistics for Data Science (DataSci w203)
% Week 1 homework
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Setup
\documentclass[12pt,a4paper]{article}
\usepackage[inner=1.5cm,outer=1.5cm,top=2.5cm,bottom=2.5cm]{geometry}
\usepackage{graphicx}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amssymb}
\numberwithin{equation}{subsection}
\usepackage{hyperref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\begin{document}

\title{Statistics for Data Science \\
       Unit 3 Homework: Probability Theory}
%\institute{Institution}
\maketitle

%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
\begin{enumerate}

\item \textbf{Gas Station Analytics}
At a certain gas station, 40\% of customers use regular gas (event R), 35\% use mid-grade (event M), and 25\% use premium (event P).  Of the customers that use regular gas, 30\% fill their tanks (Event F).  Of the customers that use mid-grade gas, 60\% fill their tanks, while of those that use premium, 50\% fill their tanks.  Assume that each customer is drawn independently from the entire pool of customers.

\begin{enumerate}
\item What is the probability that the next customer will request regular gas and fill the tank? 
\item What is the probability that the next customer will fill the tank?
\item Given that the next customer fills the tank, what is the conditional probability that they use regular gas?
\end{enumerate}

\item \textbf{The Toy Bin}

 In a collection of toys, $1/2$ are red, $1/2$ are waterproof, and $1/3$ are cool. $1/4$ are red and waterproof.  $1/6$ are red and cool. $1/6$ are waterproof and cool. $1/6$ are neither red, waterproof, nor cool. Each toy has an equal chance of being selected.
 
 \begin{enumerate}
\item Draw an area diagram to represent these events.
\item What is the probability of getting a red, waterproof, cool toy?
\item You pull out a toy at random and you observe only the color, noting that it is red.  Conditional on just this information, what is the probability that the toy is not cool?
\item Given that a randomly selected toy is red or waterproof, what is the probability that it is cool?
\end{enumerate}
 
 
\item \textbf{On the Overlap of Two Events}

Suppose for events A and B, $P(A) = 1/2$, $P(B) = 2/3$, but we have no more information about the events.
 \begin{enumerate}
\item What are the maximum and minimum possible values for $P(A \cap B)$?
\item What are the maximum and minimum possible values for $P(A|B)$?
\end{enumerate}


\item \textbf{Can't Please Everyone!}
Among Berkeley students who have completed w203, $3/4$ like statistics.  Among Berkeley students who have not completed w203, only $1/4$ like statistics.  Assume that only 1 out of 100 Berkeley students completes w203.  Given that a Berkeley student likes statistics, what is the probability that they have completed w203?


\end{enumerate}

\end{document}